Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-27 Manuscript/manuscript.Rmd
results edits

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. These associations are termed predictor-types and make up the feature space of classifiers. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap betwee a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Secondly, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as twice the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, the PI is bounded by 0 and 1. A dataset is considered balanced if PI $\approx$ 1, and imbalanced if PI$<<$1. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

@@

Class imbalance was characterized by each dataset's *proportion of imbalance* (PI) defined as the ratio of minority class observations to the sum of minority and majority class observations:

$$PI = \dfrac{C_{min}}{C_{min} + C_{maj}}$$

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the complementary set non-TAD boundary regions. Here, PI is bounded by 0 and 0.50. Heavily imbalanced classes tend to approach 0, while balanced classes approach 0.50.

@@

Once we established our optimally built and feature reduced random forest classifier, we aimed to leverage the enrichment of known annotations to predict TAD boundaries at base pair resolution. We established a *predicted TAD boundary region* as a collection of adjacent base pairs that were given a probability of 1 by our model in being a predicted TAD boundary. To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary region*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. Lastly, the midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a predicted TAD boundary.

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [??? Add ref, see one in SpectralTAD]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the five resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). A significant number [??? Need to put exactly, e.g., $18.34 \pm 5.06 \%$] of TAD boundary regions were found to be specific to one of the two cell lines (Supplementary Figure 2). Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) [??? I changed PI to CI - change in supplementary accordingly] featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The CI was found to be invariant to resolution, remaining consistently low, with an average of 0.06, 0.10, 0.14, 0.12, and 0.06 [??? Add exact numbers, like $18.34 \pm 5.06 \%$] for the Gm12878 data at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions, respectively, with similar results observed for the K562 cell line  (Supplementary Figure 3). The presence of class imbalance justifies the the need for using resampling techniques before predictive modeling.

@@

**Supplementary Figure 3. Boxplots illustrating that the cell line specific proportion of class imbalance remained severe irrespective of resolution.** Proportion of imbalance (PI) was defined as the ratio of TAD boundary regions to the sum of both TAD and non-TAD boundary regions. Balanced datasets have PI close to 0.50.

>>>>

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. These associations are termed predictor-types and make up the feature space of classifiers. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap between a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Secondly, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *class imbalance* (CI). Here, CI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, CI is bounded by 0 and 0.50. A dataset is considered balanced if CI $\approx$ 0.50, and imbalanced as CI approaches 0. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

@@

Class imbalance was (CI) defined as the ratio of minority class observations to the sum of minority and majority class observations:

$$CI = \dfrac{C_{min}}{C_{min} + C_{maj}}$$

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the complementary set non-TAD boundary regions. Here, CI is bounded by 0 and 0.50. Heavily imbalanced classes tend to approach 0, while balanced classes approach 0.50.

@@

Once we established our optimally built and feature reduced random forest classifier, we aimed to leverage the enrichment of known annotations to predict TAD boundaries at base pair resolution. We established a *predicted TAD boundary region* as a collection of adjacent base pairs that were given a probability of 1 by our model in being a predicted TAD boundary. To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary region*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. Lastly, the midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a *predicted TAD boundary*.

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin] [??? Add ref, see one in SpectralTAD; !!!added]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the five resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only 18.0% (SD=4.6%) [??? Need to put exactly, e.g., $18.34 \pm 5.06 \%$; !!!added] of TAD boundary regions overlapped between both cell lines (Supplementary Figure 2). Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) [??? I changed PI to CI - change in supplementary accordingly; !!!fixed] featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an average of 0.06, 0.10, 0.14, 0.12, and 0.06 [??? Add exact numbers, like $18.34 \pm 5.06 \%$] for the Gm12878 data at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions, respectively, with similar results observed for the K562 cell line  (Supplementary Figure 3). The presence of class imbalance justifies the the need for using resampling techniques before predictive modeling.

@@

**Supplementary Figure 3. Boxplots illustrating that the cell line specific class imbalance remained severe irrespective of resolution.** Class imbalance (CI) was defined as the ratio of TAD boundary regions to the sum of both TAD and non-TAD boundary regions. The boxplots provide summaries of CI across chromosomes across resolution, for the GM12878 (A) and K562 (B) cell lines.
